---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tdarec

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/tdarec)](https://CRAN.R-project.org/package=tdarec)
<!-- badges: end -->

The goal of {tdarec} is to provide [{recipes}](https://cran.r-project.org/package=recipes)-style preprocessing steps to compute persistent homology (PH) and calculate vectorizations of persistence data (diagrams; PDs), and to provide [{dials}](https://cran.r-project.org/package=dials)-style hyperparameter tuners to optimize these steps in ML workflows.

You can install the development version of tdarec from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("tdaverse/tdarec")
```

## Design

### Recipe steps

The current version provides two engines to compute PH (more will be implemented; see [this issue](https://github.com/corybrunson/tdarec/issues/2) for plans):

* **Vietoris--Rips** filtrations of point clouds (distance matrices or coordinate matrices) using [{ripserr}](https://github.com/tdaverse/ripserr)
* **cubical** filtrations of rasters (pixelated or voxelated data) using {ripserr}

Also included are a pre-processing step to introduce **Gaussian blur** to rasters and a post-processing step to select PDs for **specific homological degrees**.

Finally, this version provides steps that deploy the highly efficient **vectorizations** implemented in [{TDAvec}](https://github.com/uislambekov/TDAvec).
These were written with {Rcpp} specifically for ML applications.

### Tunable parameters

Most steps come with new tunable parameters, for example the maximum homological degree of the VR filtration and the number of levels in persistence landscapes.

One set of parameters that are conspicuously untunable are the "scale sequences"---the values at (or intervals over) which each transformed PD is vectorized.
An implementation is underway.

### Data formats and sets

While the most common {recipes} are designed for structured tabular data, i.e. columns with numeric or categorical entries, almost all data subjected to machine learning with persistent homology has been in forms like point clouds or greyscale images that must be stored in list-columns.
All {tdarec} examples use data in this form, and the data installed with the package is pre-processed for such use.

## Example

This example uses existing engines to optimize a simple classification model for point clouds sampled from different embeddings of the Klein bottle.
Note also that [{glmnet}](https://cran.r-project.org/package=glmnet) and [{tdaunif}](https://cran.r-project.org/package=tdaunif) must be installed.

### Setup

While not required, we attach Tidyverse and Tidymodels for convenience (with messages suppressed):

```{r example packages, message=FALSE}
# prepare a Tidymodels session and attach {tdarec}
library(tidyverse)
library(tidymodels)
library(tdarec)
```

The points are sampled uniformly from one of two Klein bottle embeddings, determined by a coin toss between the flat and the tube.

```{r example samples}
# generate samples from two embeddings
set.seed(20024L)
tibble(embedding = sample(c("flat", "tube"), size = 48, replace = TRUE)) |> 
  mutate(sample = lapply(embedding, function(emb) {
    switch(
      emb,
      flat = tdaunif::sample_klein_flat(60, sd = .5),
      tube = tdaunif::sample_klein_tube(60, sd = .5)
    )
  })) |> 
  mutate(embedding = factor(embedding)) |>
  print() -> klein_data
```

We apply a classical partition into 80% training and 20% testing sets and prepare to perform 3-fold cross-validation on the training set.

```{r example partition}
# partition the data
klein_split <- initial_split(klein_data, prop = .8)
klein_train <- training(klein_split)
klein_test <- testing(klein_split)
klein_folds <- vfold_cv(klein_train, v = 3L)
```

In this example, we adopt a common transformation of persistence diagrams, Euler characteristic curves.
For their vectorization, we need a scale sequence that spans the birth and death times of any persistent features, and for this we choose a round number larger than the diameters of both point clouds (based on the sampler documentation) as an upper bound.
Rather than choose _a priori_ to use homology up to degree 0, 1, 2, or 3, we prepare to tune the maximum degree during optimization.

### Specifications

```{r example recipe}
# specify a pre-processing recipe
scale_seq <- seq(0, 3, by = .05)
recipe(embedding ~ sample, data = klein_train) |> 
  step_phom_point_cloud(
    sample, max_hom_degree = tune("vr_degree"),
    keep_original_cols = FALSE
  ) |> 
  step_vpd_euler_characteristic_curve(
    sample_phom, xseq = scale_seq,
    keep_original_cols = FALSE
  ) |> 
  print() -> klein_rec
```

For simplicity, we choose a common model for ML classification, penalized logistic regression.
We fix the mixture coefficient to use LASSO rather than ridge regression but prepare the penalty parameter for tuning. 

```{r example model}
# specify a classification model
logistic_reg(penalty = tune(), mixture = 1) |> 
  set_mode("classification") |> 
  set_engine("glmnet") |> 
  print() -> klein_lm
```

We then generate a complete hyperparameter tuning grid by crossing the grids generated for the two unspecified parameters:

```{r example grid}
# generate a hyperparameter tuning grid
klein_rec_grid <- grid_regular(
  extract_parameter_set_dials(klein_rec), levels = 3,
  filter = c(vr_degree > 0)
)
klein_lm_grid <- grid_regular(
  extract_parameter_set_dials(klein_lm), levels = 5
)
klein_grid <- merge(klein_rec_grid, klein_lm_grid)
```

### Optimization

We evaluate the model across the hyperparameter grid using cross-validation, using the area under the sensitivity--specificity (ROC) curve:

```{r example tune}
# optimize the model performance
klein_res <- tune_grid(
  klein_lm,
  preprocessor = klein_rec,
  resamples = klein_folds,
  grid = klein_grid,
  metrics = metric_set(roc_auc)
)
```

From the results, we obtain the best-performing parameter setting:

```{r example metric}
klein_res |> 
  select_best(metric = "roc_auc") |> 
  print() -> klein_best
```

### Evaluation

This optimal setting includes both the VR homology degree and the GLM penalty, so both the pre-processing recipe and the predictive model must be finalized in order to fit the final model to the full training set:

```{r example fit}
klein_rec_fin <- klein_rec |> finalize_recipe(klein_best) |> prep()
klein_lm_fin <- klein_lm |> finalize_model(klein_best)
klein_fit <- fit(
  klein_lm_fin,
  formula(klein_rec_fin),
  data = bake(klein_rec_fin, new_data = klein_train)
)
```

Finally, we evaluate the fitted model on the testing set:

```{r example evaluate}
klein_fit |> 
  predict(
    new_data = bake(klein_rec_fin, new_data = klein_test),
    type = "prob"
  ) |> 
  bind_cols(select(klein_test, embedding)) |> 
  roc_auc(truth = embedding, .pred_flat)
```

## Contributions

Please note that the tdarec project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

### Generated code

Much of the code exposing {TDAvec} tools to Tidymodels is generated by elaborate scripts rather than written manually.
While maintenance of these scripts takes effort, it prevents (or at least flags) errors arising from cascading implications of changes to the original functions, and it allows simple and rapid package-wide adjustments. If you see an issue with generated code, please raise an issue to discuss it before submitting a pull request.

### Acknowledgments

This project was funded by [an ISC grant from the R Consortium](https://r-consortium.org/all-projects/2024-group-1.html#modular-interoperable-and-extensible-topological-data-analysis-in-r) and done in coordination with Aymeric Stamm and with guidance from Bertrand Michel and Paul Rosen.
It builds upon the work of and conversations with Umar Islambekov and Aleksei Luchinsky, authors of [{TDAvec}](https://github.com/uislambekov/TDAvec).
Package development also benefitted from the support of colleagues in [the Laboratory for Systems Medicine](https://systemsmedicine.pulmonary.medicine.ufl.edu/) and [the TDA Seminar](https://tda.math.ufl.edu/) and the use of equipment at [the University of Florida](https://www.ufl.edu/).
