---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tdarec

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of {tdarec} is to provide additional preprocessing steps to [{recipes}](https://github.com/tidymodels/recipes) to compute persistent homology (PH) and calculate vectorizations of persistence data (diagrams; PDs).

The current prototype provides one engine to compute PH:

* Vietoris--Rips filtrations of point clouds using [{ripserr}](https://github.com/tdaverse/ripserr)

and one engine to vectorize PDs:

* Euler characteristic curves using [{TDAvec}](https://github.com/uislambekov/TDAvec).

The goal is to provide all PH and PD vectorization engines published on CRAN.

## Installation

You can install the development version of tdarec from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("corybrunson/tdarec")
```

## Example

This example uses existing engines in a full Tidyverse workflow to optimize a simple classification model for point clouds sampled from different embeddings of the Klein bottle:

```{r example}
# prepare a Tidymodels session and attach {tdarec}
library(tidyverse)
library(tidymodels)
library(tdarec)

# generate samples from two embeddings
set.seed(20024L)
tibble(embedding = sample(c("flat", "tube"), size = 48, replace = TRUE)) |> 
  mutate(sample = lapply(embedding, function(emb) {
    switch(
      emb,
      flat = tdaunif::sample_klein_flat(60, sd = .5),
      tube = tdaunif::sample_klein_tube(60, sd = .5)
    )
  })) |> 
  # mutate(embedding = factor(embedding)) |> 
  print() -> klein_data

# partition the data
klein_split <- initial_split(klein_data, prop = .8)
klein_train <- training(klein_split)
klein_test <- testing(klein_split)
klein_folds <- vfold_cv(klein_train, v = 3L)

# specify a pre-processing recipe
scale_seq <- seq(0, 3, by = .05)
recipe(embedding ~ sample, data = klein_train) |> 
  step_phom_point_cloud(
    sample, max_hom_degree = tune("vr_degree"),
    keep_original_cols = FALSE
  ) |> 
  step_vpd_euler_characteristic_curve(
    sample_phom, xseq = scale_seq,
    keep_original_cols = FALSE
  ) |> 
  print() -> klein_rec

# specify a classification model
logistic_reg(penalty = tune(), mixture = 1) |> 
  set_mode("classification") |> 
  set_engine("glmnet") |> 
  print() -> klein_lm

# generate a hyperparameter tuning grid
klein_rec_grid <- grid_regular(
  extract_parameter_set_dials(klein_rec), levels = 3,
  filter = c(vr_degree > 0)
)
klein_lm_grid <- grid_regular(
  extract_parameter_set_dials(klein_lm), levels = 5
)
klein_grid <- merge(klein_rec_grid, klein_lm_grid)

# optimize the model performance
klein_res <- tune_grid(
  klein_lm,
  preprocessor = klein_rec,
  resamples = klein_folds,
  grid = klein_grid,
  metrics = metric_set(roc_auc, pr_auc)
)
klein_res |> 
  collect_metrics()
klein_res |> 
  select_best(metric = "roc_auc")
```

## Code of Conduct

Please note that the tdarec project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
